---
title: "ggplot2-theme"
author: "Pawan"
date: "2022-12-08"
output: html_document
---

# Themes in ggplot2

The ggplot2 package have different themes. By default it uses the theme_grey (```theme_gray()```), so we don’t need to specify it.

We can adjusted themes by adding an in-build theme from ggplot2.

```theme_gray()``` – default ggplot2 theme  
```theme_bw()``` – dark and light (black and white) ggplot2 theme  
```theme_linedraw()``` – uses black lines on white backgrounds only  
```theme_light()``` – similar to linedraw() but with grey lines as well  
```theme_dark()``` – lines on a dark background instead of light  
```theme_minimal()``` – no background annotations, minimal feel  
```theme_classic()``` – theme with no grid lines  
```theme_void()``` – empty theme with no elements  



```{r}
library(ggplot2)

mtcars->dt
cyls <- as.factor(dt$cyl)

ggplot(dt, aes(x = mpg, fill = cyls)) +
  geom_density(alpha = 0.7)->p

## theme_bw()
p+theme_bw() +
  theme(legend.position = "top")->p1 

## theme_dark()
p+theme_dark() +
  theme(legend.position = "top")->p2 

p1
p2
```

# ggthemes package

The <mark>ggthemes</mark> package contains several very popular themes.

```
theme_base()         theme_cala()                  theme_clean()
theme_economist()    theme_excel()                 theme_excel_new()
theme_few()          theme_fivethirtyeight()       theme_foundation()
theme_gdocs()        theme_hc()                    theme_igray()
theme_map()          theme_pander()                theme_par()
theme_solarized()    theme_solid()                 theme_stata()
```

```{r}
library(ggplot2)
## install.packages("ggthemes")
library(ggthemes)

mtcars->dt
cyls <- as.factor(dt$cyl)

ggplot(dt, aes(x = mpg, fill = cyls)) +
  geom_density(alpha = 0.7)->p

## theme_base()
p+theme_base() +
  theme(legend.position = "top")->p1

## theme_economist()
p+theme_economist() +
  theme(legend.position = "top")->p2
p1
p2


```

                 

# hrbrthemes package

<mark>hrbrthemes</mark> provides “typography-centric themes and theme components for ggplot2”. The available themes are listed below, but you can also check the available scales, palettes, fonts and utilities.

```
theme_ipsum()              theme_ipsum_es()             theme_ipsum_rc()
theme_ipsum_ps()           theme_ipsum_pub()            theme_ipsum_tw()
theme_ipsum_modern_rc()    theme_ft_rc()
```

       
```{r}
library(ggplot2)
## install.apckages("hrbrthemes")
library(hrbrthemes)

mtcars->dt
cyls <- as.factor(dt$cyl)

ggplot(dt, aes(x = mpg, fill = cyls)) +
  geom_density(alpha = 0.7) + 
  ggtitle("Plot title") +
  theme_ipsum() + # Arial Narrow
  scale_fill_ipsum() +
  theme(legend.position = "top") 
```
 
# ggthemr package

Note that you will need to call ggthemr_reset() to reset to the default ggplot2 theme.

The ggthemr package works different than the other packages. Instead of using the theme function and set a theme you can set a theme globally making use of the ggthemr function and passing the theme as string as argument.

```
ggthemr("flat")           ggthemr("flat dark")          ggthemr("camouflage")
ggthemr("carrot")         ggthemr("chalk")              ggthemr("copper")
ggthemr("dust")           ggthemr("earth")              ggthemr("fresh")
ggthemr("grape")          ggthemr("grass")              ggthemr("greyscale")
ggthemr("light")          ggthemr("lilac")              ggthemr("pale")
ggthemr("sea")            ggthemr("sky")                ggthemr("solarized")
```


```{r}
library(ggplot2)
##devtools::install_github('Mikata-Project/ggthemr')
library(ggthemr)

mtcars->dt
cyls <- as.factor(dt$cyl)

ggthemr("flat")

ggplot(dt, aes(x = mpg, fill = cyls)) +
  geom_density(alpha = 0.7) +
  theme(legend.position = "top") 
```

                 

# ggtech package
The <mark>ggtech</mark> package provides themes inspired by tech companies, such as Airbnb, Google, Twitter or Facebook.

```
theme_tech(theme="airbnb")               theme_tech(theme="etsy")
theme_tech(theme="facebook")             theme_tech(theme="google")
theme_tech(theme="tiwtter")              theme_airbnb_fancy()
```

```{r}
library(ggplot2)
##devtools::install_github("ricardo-bion/ggtech", 
##                          dependencies=TRUE)
library(ggtech)

mtcars->dt
cyls <- as.factor(dt$cyl)

ggplot(dt, aes(x = mpg, fill = cyls)) +
  geom_density(alpha = 0.7) +
  ggtitle("Title of the plot") + 
  theme_tech(theme = "airbnb") +
  scale_fill_tech(theme = "airbnb") +
  theme(legend.position = "top") 

ggplot(dt, aes(x = mpg, fill = cyls)) +
  geom_density(alpha = 0.7) +
  ggtitle("Title of the plot") + 
  theme_tech(theme = "facebook") +
  scale_fill_tech(theme = "facebook") +
  theme(legend.position = "top") 
```

# ggdark package

ggdark provides the dark mode themes of the in-built ggplot2 themes. In addition, the package can convert any theme into a dark theme making use of the dark_mode function.

```
dark_theme_gray()         dark_theme_linedraw()          dark_theme_light()          dark_theme_dark()         dark_theme_minimal()           dark_theme_classic()        dark_theme_void()         dark_mode()
```

```{r}
library(ggplot2)
## install.packages("ggdark")
library(ggdark)

mtcars->dt
cyls <- as.factor(dt$cyl)

ggplot(mtcars, aes(x = mpg, fill = cyls)) +
  geom_density(alpha = 0.7) +
  dark_theme_gray() +
  theme(legend.position = "top") 
```


# Set an active theme
The theme_set function can be used to set a theme globally. In the following example we are setting the in-built theme_dark to all the generated plots. Then, you can you back to the default theme as follows.

```{r}
library(ggplot2)

# Global theme
old <- theme_set(theme_dark())

cyls <- as.factor(mtcars$cyl)

ggplot(mtcars, aes(x = mpg, fill = cyls)) +
  geom_density(alpha = 0.7) +
  theme(legend.position = "top")

# Reset to default
theme_set(old) 
``` 