---
title: "ggplot-theme"
author: "Pawan"
date: "2022-12-08"
output: pdf_document
---

# Themes in ggplot2
In-built themes
ggthemes package
hrbrthemes package
ggthemr package
ggtech package
ggdark package
Set an active theme
You can customize the appearance of your plots made with ggplot2 using themes created by other users. Click on the buttons of each section to visualize each theme and its corresponding code.

Note that you can override theme elements making use of the theme function, like the background color, the grid lines or the margins, among others.

In-built themes
The ggplot2 package comes with eight different themes. By default it uses the theme named theme_grey (theme_gray), so you don’t really need to specify it.

Choose a theme

       

Note that there is an additional theme named theme_test to conduct visual unit tests by advanced users.

Default ggplot2 theme

```{r}
library(ggplot2)

cyls <- as.factor(mtcars$cyl)

ggplot(mtcars, aes(x = mpg, fill = cyls)) +
  geom_density(alpha = 0.7) +
  theme_grey() +  # Default
  theme(legend.position = "top") 
```

ggthemes package

```{r}
library(ggplot2)
library(ggthemes)

cyls <- as.factor(mtcars$cyl)

ggplot(mtcars, aes(x = mpg, fill = cyls)) +
  geom_density(alpha = 0.7) +
  theme_base() +
  theme(legend.position = "top") 
```

The ggthemes package contains several very popular themes. Some of them also come with their corresponding color scales. Use the scales properly according to your data.

                 

hrbrthemes package
hrbrthemes provides “typography-centric themes and theme components for ggplot2”. The available themes are listed below, but you can also check the available scales, palettes, fonts and utilities.

       
```{r}
library(ggplot2)
library(hrbrthemes)

cyls <- as.factor(mtcars$cyl)

ggplot(mtcars, aes(x = mpg, fill = cyls)) +
  geom_density(alpha = 0.7) + 
  ggtitle("Plot title") +
  theme_ipsum() + # Arial Narrow
  scale_fill_ipsum() +
  theme(legend.position = "top") 
```
 
ggthemr package

```{r}
library(ggplot2)
library(ggthemr)

cyls <- as.factor(mtcars$cyl)

ggthemr("flat")

ggplot(mtcars, aes(x = mpg, fill = cyls)) +
  geom_density(alpha = 0.7) +
  theme(legend.position = "top") 
```

Note that you will need to call ggthemr_reset() to reset to the default ggplot2 theme.

The ggthemr package works different than the other packages. Instead of using the theme function and set a theme you can set a theme globally making use of the ggthemr function and passing the theme as string as argument.

                 

ggtech package
The ggtech package provides themes inspired by tech companies, such as Airbnb, Google, Twitter or Facebook.

```{r}
library(ggplot2)
library(ggtech)

cyls <- as.factor(mtcars$cyl)

ggplot(mtcars, aes(x = mpg, fill = cyls)) +
  geom_density(alpha = 0.7) +
  ggtitle("Title of the plot") + 
  theme_tech(theme = "airbnb") +
  scale_fill_tech(theme = "airbnb") +
  theme(legend.position = "top") 
```

ggdark package

```{r}
library(ggplot2)
library(ggdark)

cyls <- as.factor(mtcars$cyl)

ggplot(mtcars, aes(x = mpg, fill = cyls)) +
  geom_density(alpha = 0.7) +
  dark_theme_gray() +  # Default
  theme(legend.position = "top") 
```

ggdark provides the dark mode themes of the in-built ggplot2 themes. In addition, the package can convert any theme into a dark theme making use of the dark_mode function.

        

Set an active theme
The theme_set function can be used to set a theme globally. In the following example we are setting the in-built theme_dark to all the generated plots. Then, you can you back to the default theme as follows.

```{r}
library(ggplot2)

# Global theme
old <- theme_set(theme_dark())

cyls <- as.factor(mtcars$cyl)

ggplot(mtcars, aes(x = mpg, fill = cyls)) +
  geom_density(alpha = 0.7) +
  theme(legend.position = "top")

# Reset to default
theme_set(old) 
``` 

Set an active theme in ggplot2 with the theme_set function
