---
title: "ggplot2"
author: "Pawan"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

ggplot2 is an open-source of data visualization and graphical representation package for the R programming language. This package is created by [Hadley Wickham](https://hadley.nz/).

# Installation
We can install ggplot2 just by using this command

```
install.packages("ggplot2")
```
Note: ```ggplot2``` is also included in the ```tidyverse``` package.

Loading the ggplot2 package
```{r}
library(ggplot2)
```
# Let's start plotting with ggplot2
There are three basic parts of ggplot   
i) First we to define the data from which we want to plot our graph  

ii) mapping, here we can define our aesthetics like x-axis, y-axis, color, linetpye, shape, etc.
```aes(x, y, color)```  

iii) geom function, geom function define the type of graph that we want to plot, like scatter plot, line plot, bar plot, box plot, etc.
``` geom_line()```, ```geom_point()```  


```ggplot(data, aes(x, y)) + geom_function()```

Let's start to plot a simple graph

```{r}
library(ggplot2)

iris->dataset

## Scatter plot
ggplot(dataset, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point() ->p1

p1
## Bar plot
ggplot(dataset, aes(x = Sepal.Length, y = Sepal.Width, fill = Species)) +
  geom_bar(stat = "identity", position = "stack") ->p2

p2
```

ggplot2 provide many geom function as shown below:  

`geom_point()` Scatter Plot  

`geom_line()`  Line plot

`geom_bar()`   Bar charts

`geom_blank()` Draw a blank plot

`geom_boxplot()`  Box plot   

`geom_density()`  Density plot

`geom_dotplot()` Dot plot

`geom_histogram()` Histograms

`geom_map()`     Map plot

`geom_polygon()`  Polygons plot

`geom_ribbon()` or `geom_area()`  Ribbons and area plots

`geom_segment()` or `geom_curve()`  Line segments and curves

`geom_smooth()` or `stat_smooth()` Plot Conditional means

`geom_label()` or `geom_text()`  Text plot

`geom_violin()`  Violin plot


# Facet 


```{r}
library(ggplot2)
iris->dataset

ggplot(dataset, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point() ->p

p+ facet_wrap(~Species)->p
p

```

# Theme
The ggplot2 package have different themes. By default it uses the theme_grey (```theme_gray()```), so we don’t need to specify it. Themes provided by ggplot are below


```theme_gray()``` – default ggplot2 theme with a grey background and white gridlines   
```theme_bw()``` – dark and light (black and white) ggplot2 theme    
```theme_linedraw()``` – uses black lines on white backgrounds only    
```theme_light()``` – similar to linedraw() but with grey lines    
```theme_dark()``` – lines on a dark background    
```theme_minimal()``` – no background annotations    
```theme_classic()``` – theme with no grid lines  
```theme_void()``` – empty theme with no elements    

We can also modify the plot base size by using <mark>base_size</mark> argument in the theme functions. 

```{r}
library(ggplot2)

iris->dataset

# Scatter Plot
ggplot(dataset, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point() ->p

p+ facet_wrap(~Species)->p

## theme_bw()
p+theme_bw(base_size=20)->p1
p1

## theme_dark()
p+theme_dark()->p2
p2
```

# Lab funtion
Whit using of ```labs()``` function we can set a title, subtitle, caption, and axis title of the plot.

```{r}
library(ggplot2)

iris->dataset

# Scatter Plot
ggplot(dataset, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point() ->p

p+ facet_wrap(~Species)->p

## theme_bw()
p+theme_bw(base_size=20)->p

## Lab Function
p+labs(title = "Title of the plot",
       subtitle = "Subtitle of the plot",
       caption = "Caption of the plot",
       x = "x-axis",
       y = "y-axis")->p
p
```

# Modify plot with theme function
We can modify our plot with the use of ```theme()`` function. Here we can define position of the title, subtile, caption, panel background, legend position, and many more things. You can define your own theme, size, color, shape of the plot as you want. Below given some use theme arguments which are also my favorite

```
theme(
  axis.title,        ## for both axis
  axis.title.x,      ## for x axis
  axis.title.y,      ## for y axis
  axis.text,         ## for both axis
  axis.text.x,       ## for x axis
  axis.text.y,       ## for y axis
  axis.ticks,        ## for both axis
  axis.ticks.x,      ## for x axis
  axis.ticks.y,      ## for y axis
  axis.line,         ## for both axis
  axis.line.x,       ## for x axis
  axis.line.y,       ## for y axis
  legend.background,
  legend.margin,
  legend.text,
  legend.title,
  legend.position,
  legend.direction,
  legend.box,
  panel.background,
  panel.border,
  panel.grid,
  panel.grid.major,
  panel.grid.minor,
  plot.background,
  plot.title,
  plot.subtitle,
  plot.caption,
  plot.tag,
  plot.margin,
  strip.background,
  strip.background.x,
  strip.background.y,
  strip.text,
  strip.text.x,
  strip.text.y)
```

With the using of these arguments we are able to modify plot, these arguments   
```element_line()``` use for line  
```element_rect()``` use for color  
```element_text()``` use for text  
```element_blank()``` use for blank 


For example 
```
theme(plot.title = element_text(hjust = 1),
      plot.subtitle = element_text(size = 0.5),
      panel.background = element_rect(fill = "yellow"),
      axis.text = element_text(size = 2))
```        
        

```{r}
library(ggplot2)

iris->dataset

# Scatter Plot
ggplot(dataset, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point() ->p

p+ facet_wrap(~Species)->p

## theme of the plot
p+theme_bw()->p

## Lab Function
p+labs(title = "Title of the plot",
       subtitle = "Subtitle of the plot",
       caption = "Caption of the plot")->p

## theme function
p+theme(legend.position = "bottom",
        plot.title = element_text(family = "serif",              # Font family
                                  face = "bold",                 # Font face
                                  size = 15,                     # Font size
                                  hjust = 1,                     # Horizontal adjustment
                                  vjust = 1),                   # Vertical adjustment
        plot.subtitle = element_text(hjust = 1),
        plot.caption = element_text(hjust = 0.1),
        panel.background = element_rect(fill = "yellow"),
        axis.text = element_text(size = 2),
        strip.background = element_rect(fill = "pink"),
        strip.text = element_text(size = 9),
        plot.background = element_rect(fill = "green"),
        panel.border = element_rect(linetype = "dashed"),
        panel.grid.major = element_line(colour = "black"))->p 
p
```

# Combine multiple plot 

The ggplot2 package doesn’t providing function to arrange multiples plots in a single figure, but there are some packages, that allow combine multiples plots into a single figure with custom layouts, width, height, such as <mark>cowplot</mark>, <mark>gridExtra</mark>, <mark>patchwork</mark>.
(https://medium.com/@pawanjangra1198/combining-plots-in-ggplot2-9699acaa2942#6632-36f5b0859cff)


```{r}
library(ggplot2)

iris->dataset

# Scatter Plot
ggplot(dataset, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point() ->p

p+ facet_wrap(~Species)->p

## theme_bw()
p+theme_bw(base_size=20)->p1
p1

## theme_dark()
p+theme_dark()->p2
p2
```


The <mark>gridExtra</mark> package provides the <mark>grid.arrange</mark> function to combine several plots on a single figure.  

```{r}
# install.packages("gridExtra")
library(ggplot2)
library(gridExtra)

# Combine the plots with gridExtra
grid.arrange(p1, p2) 
```


# Save plot
We can save the plot by using `ggsave()` function and in this we can define the height, width of the plot.

```{r}
library(ggplot2)

iris->dataset

# Scatter Plot
ggplot(dataset, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point() ->p

p+ facet_wrap(~Species)->p

## theme_bw()
p+theme_bw(base_size=20)->p

## save the plot
ggsave("~/ggplot/plot.png", width = 3200, height = 2800, unit = "px")


```